Going Beyond Ideal Point Points: Modeling Measurement Model Measurement Error

### 2024-04-09 paper.qmd
diss draft commit

@@

In @fig-measurement-model, an estimate from distribution B should be treated as more uncertain than one from distribution A when used to test a theory. Failing to do so, as I will show, can lead to attenuation bias---or the false conclusion that the latent variable has no association with an outcome when it in fact does. In other words, the method I propose can help increase the statistical power of theory tests. Panels C and D in @fig-measurement-model show skewed distributions. Here the danger is that the skewness is caused by a third variable, which *also* causes the outcome of interest in the theory-testing model. This, as I will show, can lead to confounding bias if the skewness of the measurement output is not accounted for.

@@

There are two practical issues, however, with building a fully-specified joint measurement and theory-testing model. The first is computational. Bayesian statistical software uses notoriously expensive Markov Chain Monte Carlo (MCMC) sampling methods to derive its posterior distributions. Even on their own, measurement models which use MCMC can be extremely slow to sample given these types of models' high-dimensional nature. So attempting to sample from a model which also includes an arbitrarily complex theory-testing model, $f(\cdot)$, in addition to the measurement model may simply be unfeasible given the computing power that the average researcher has access to. The second challenge with the idealized joint model is that it requires researchers to write down a fully-specified measurement model, $g(\cdot)$. Compared to their theory-testing model, applied researchers likely have much less knowledge regarding the intricacies involved in estimating latent variables. Because latent variables have no objective scale, measurement models can often be challenging to fit in practice due to issues of identifiability.

@@

Let's say we want to estimate the effect of legislator ideology, $\theta$, on some outcome, $y$. Using the format developed in @eq-ex-equation, @eq-irt shows an example joint measurement and theory-testing model to answer this question using ideology estimates from an IRT model. The measurement model (bottom) predicts whether a legislator voted yes or no on a piece of legislation, $z_j$, using the traditional 2-parameter IRT equation $\Phi(\gamma_j\theta_i + \xi_j)$. Estimates of the parameters $\theta_i$ from this model, are then used as data in the linear regression theory-testing model (top) to estimate $\beta_1$---the coefficient of substantive interest.

$$
\begin{aligned}
  y_i &\sim \text{Normal}(\beta_0 + \beta_1 \theta_i, \sigma^2) \\
  z_{ij} &\sim \text{Bernoulli}[\Phi(\gamma_j\theta_i + \xi_i)] \\
\end{aligned}
$$ {#eq-irt}

@@

Using the logic of causal graphs, I will discuss two types of measurement model measurement error, and how the joint measurement theory-testing procedure laid out in @eq-ex-equation helps fix them. First I will consider random, or classical, measurement error. In this case the joint model will (in most cases) provide researchers with extra statistical power to test their theory by mitigating issues of attenuation bias. Then we will look at non-random measurement error scenarios, in which the measurement model error introduces confounding in the theory-testing model. I will show how the joint model from @eq-ex-equation ameliorates this confounding bias. For both types of measurement error, I will use simulation studies to demonstrate how effective each modeling approach is at recovering known parameter values from the theory-testing model.

@@

If the posterior distributions for the latent parameters $\theta_i$ from the measurement model follow a normal distribution, this is a form of classical measurement error. Here we do not assume that there is some relationship between the measurement error and the outcome of interest, rather, the errors are simply random "noise" in the measurement estimates. Classical measurement error leads to attenuation bias: a reduction of the main effect size in the theory-testing model towards zero. Thus, the wider the posterior distribution is for $\bar{\theta}_i$, the more likely we are to make a False Negative error in our theory-testing model.

Let's return to the Bayesian IRT measurement model used to estimate political ideology for legislators. @fig-ideal-dag shows the causal process which produces these ideal point estimates. The observed measurements of ideology, $\bar{\theta}$ come from an unobserved latent variable plus some measurement error. At least two variables can affect the measurement error, $e^{\theta}$ in @fig-ideal-dag. First, true ideal point of the group, $\theta$ influences the amount of measurement error for estimates of the group's ideology because, as we move further from the ideological center of the scale, uncertainty increases. @fig-irt-dist shows an example of what the distribution of posterior estimates from a Bayesian IRT model look like. Groups further from the center have much wider ideal point posterior distributions. The second variable affecting $e^{\theta}$ is participation, $P$---how much a particular legislator has taken positions on bills. Legislators that signal more positions on bills will have smaller levels of measurement error compared to those that signal fewer positions because we have more data on their true ideological preferences.

@@

There is also likely unobserved measurement error in these types of models. IRT models assume each legislator's decision to vote on bill is influenced solely by their inner ideology, rather than on strategic concerns. A violation of this assumption, therefore, will produce biased estimates of $\bar{\theta}$. There are also computational issues with fitting IRT models which can make the posterior estimates untrustworthy. For the purposes of this illustration, however, I will assume that the posterior distribution, $e^{\theta}$ for $\bar{\theta}$ contains all relevant information about the measurement error for the true ideology $\theta$.

@@

Building off from @fig-ideal-dag-theory, let's now consider the hypothetical causal graph shown in @fig-ideal-dag-theory-c. As before, the main causal effect of interest is represented by the path $\bar{\theta} \longrightarrow Y$. In order to get an unbiased estimate of this causal effect we need to close all backdoor paths leading from $\bar{\theta}$ to $Y$, which in this case, flows through the unobserved variable $U$. This confound represents anything that is a common cause of both the IRT model measurement error and the outcome of interest.

@@

For the theory-testing model in @fig-ideal-dag-theory-c it may be possible to directly condition on some variables in $U$ in order to obtain an unbiased estimate of $\bar{\theta} \longrightarrow Y$. But with something as multi-faceted as political ideology, there is always some risk of residual confounding. My proposed method of building a joint measurement and theory-testing model fixes this issue by obviating the need to deal with $U$ at all. This is because the measurement error, $e^\theta$ in @fig-ideal-dag-theory-c is part of the backdoor path from $\bar{\theta}$ to $Y$. Therefore when we explicitly incorporate $e^\theta$ into a model estimating $\bar{\theta} \longrightarrow Y$ we can obtain an unbiased---or at least less-biased, given unobserved measurement error---estimate of the causal effect of ideology on the theoretical outcome of interest.

@@

After the data are generated, we fit two models: the ordinary linear regression using only $\bar{\theta}$ values as used previously in the attenuation bias example (@eq-cont-no-me-model), and a modified version of @eq-cont-me-model which substitutes the Normal distribution for the Skew Normal distribution (@eq-cont-skew-model). As before, the key parameter of theoretical interest is $\beta_1$---which, according to the simulated data, should equal zero if unconfounded.

@@

I also split the data by party to make the main effects easier to interpret. Lower values from the IRT ideology measurement model correspond to more left-wing candidates, whereas higher values correspond to more right-wing candidates. So, for Democrats we interpret a positive relationship between ideology and vote share as district electorates favoring moderate candidates, whereas a positive relationship for Republican candidates would mean district electorates favor more extremists.

@@

@fig-ree-betas shows the posterior distributions for $\beta_1$ in the above models. For Democrats (top panel), there does not appear to be a strong relationship between incumbent ideology and their electoral success. Both the Normal measurement error model and mean-only model coefficients are centered around zero. Using the Skew-Normal measurement error model, however, there appears to be some evidence that centrist Democrats perform worse electorally compared to more liberal Democrats. For Republicans (bottom panel), all models agree that more extremist candidates perform better. The fact that both measurement error models predict larger electoral gains from moving rightward suggests that there may be some attenuation bias taking place in the mean-only model.

@@

@@

In the same vein as using the best computational sampling methods, this project highlights why latent variables whose measurement models produce rich uncertainty estimates should be preferred over those that do not. The methodological competitor to IRT models for measuring political ideology is DW-NOMINATE, which uses multidimensional scaling rather than Bayesian estimation. This optimization method does not provide explicit estimates of uncertainty, much less a full posterior distribution of plausible values the latent variable could take. All of DW-NOMINATE's uncertainty estimates must come from bootstrap procedures which produce, at best, only standard errors. This means that more of this model's measurement error is in the unobserved category, which in turn raises concerns about attenuation bias and/or confounding bias when using DW-NOMINATE values in a theory-testing model.

>>>>

## The Problem

@@

In @fig-measurement-model, an estimate from distribution B should be treated as more uncertain than one from distribution A when used to test a theory. Failing to do so, as I will show, can lead to attenuation bias---or the false conclusion that the latent variable has no association with an outcome when it in fact does. In other words, the method I propose can help increase the statistical power of theory tests by reducing the rate of false negatives. Panels C and D in @fig-measurement-model show skewed distributions. Here the danger is that the skewness is caused by a third variable, which *also* causes the outcome of interest in the theory-testing model. This, as I will show, can lead to confounding bias if the skewness of the measurement output is not accounted for.

@@

There are two practical issues, however, with building a fully-specified joint measurement and theory-testing model. The first is computational. Bayesian statistical software uses notoriously expensive Markov Chain Monte Carlo (MCMC) sampling methods to derive its posterior distributions. Even on their own, measurement models which use MCMC can be extremely slow to sample given these types of models' high-dimensional nature. So attempting to sample from a model which also includes an arbitrarily complex theory-testing model, $f(\cdot)$, in addition to the measurement model may simply be unfeasible given the computing power that the average researcher has access to. The second challenge with the idealized joint model is that it requires researchers to write down a fully-specified measurement model, $g(\cdot)$. Compared to their theory-testing model, applied researchers likely have much less knowledge regarding the intricacies involved in estimating latent variables. Because latent variables have no objective scale, measurement models can often be challenging to fit in practice due to issues of model identification.

@@

Let's say we want to estimate the effect of legislator ideology, $\theta$ on some outcome, $y$. Using the format developed in @eq-ex-equation, @eq-irt shows an example joint measurement and theory-testing model to answer this question using ideology estimates from an IRT model. The measurement model (bottom) predicts whether a legislator voted yes or no on a piece of legislation, $z_j$ using the traditional 2-parameter IRT equation $\Phi(\gamma_j\theta_i + \xi_j)$. Estimates of the parameters $\theta_i$ from this model, are then treated like data in the linear regression theory-testing model (top) to estimate $\beta_1$---the coefficient of substantive interest.

$$
\begin{aligned}
  y_i &\sim \text{Normal}(\beta_0 + \beta_1 \theta_i, \sigma^2) \\
  z_{ij} &\sim \text{Bernoulli}[\Phi(\gamma_j\theta_i + \xi_j)] \\
\end{aligned}
$$ {#eq-irt}

@@

Using the logic of causal graphs, I will discuss two types of measurement model measurement error, and how the joint measurement theory-testing procedure laid out in @eq-ex-equation helps fix them. First I consider random, or classical, measurement error. In this case the joint model will (in most cases) provide researchers with extra statistical power to test their theory by mitigating issues of attenuation bias and helping avoid false negatives. Then we will look at non-random measurement error scenarios, in which the measurement model error introduces confounding in the theory-testing model. I will show how the joint model from @eq-ex-equation ameliorates this confounding bias. For both types of measurement error, I use simulation studies to demonstrate how effective each modeling approach is at recovering known parameter values from the theory-testing model.

@@

If the posterior distributions for the latent parameters $\theta_i$ from the measurement model follow a normal distribution, this is a form of classical measurement error. Here we do not assume that there is some relationship between the measurement error and the outcome of interest, rather, the errors are simply random "noise" in the measurement estimates. Classical measurement error leads to attenuation bias: a reduction of the main effect size in the theory-testing model towards zero. Thus, the wider the posterior distribution is for $\bar{\theta}_i$, the more likely we are to make a false negative error in our theory-testing model.

Let's return to the Bayesian IRT measurement model used to estimate political ideology for legislators. @fig-ideal-dag shows the causal process which produces these ideal point estimates. The observed measurements of ideology, $\bar{\theta}$ come from an unobserved latent variable plus some measurement error. At least two variables can affect the measurement error, $e^{\theta}$ in @fig-ideal-dag. First, the true ideal point of the group, $\theta$ influences the amount of measurement error for estimates of the group's ideology because, as we move further from the ideological center of the scale, uncertainty increases. @fig-irt-dist shows a stylized example of what the distribution of posterior estimates from a Bayesian IRT model can look like. Groups further from the center have increasingly wider ideal point posterior distributions. The second variable affecting $e^{\theta}$ is participation, $P$---how much a particular legislator has taken positions on bills. Legislators that signal more positions on bills will have smaller levels of measurement error compared to those that signal fewer positions because we have more information on their true ideological preferences.

@@

There is also likely unobserved measurement error in these types of models. IRT models assume that each legislator's decision to vote on bill is influenced solely by their inherent ideology, rather than by strategic concerns. A violation of this assumption, therefore, will produce biased estimates of $\bar{\theta}$. There are also computational issues with fitting IRT models which can make the posterior estimates untrustworthy. For the purposes of this illustration, however, I will assume that the posterior distribution, $e^{\theta}$ for $\bar{\theta}$ contains all relevant information about the measurement error for the true ideology $\theta$.

@@

Building off from the DAG in @fig-ideal-dag-theory, let's now consider the hypothetical causal graph shown in @fig-ideal-dag-theory-c. As before, the main causal effect of interest is represented by the path $\bar{\theta} \longrightarrow Y$. In order to get an unbiased estimate of this causal effect we need to close all backdoor paths leading from $\bar{\theta}$ to $Y$, which in this case, flows through the unobserved variable $U$. This confounding variable represents anything that is a common cause of both the IRT model measurement error and the outcome of interest.

@@

For the theory-testing model in @fig-ideal-dag-theory-c it may be possible to directly condition on some variables in $U$ in order to obtain an unbiased estimate of $\bar{\theta} \longrightarrow Y$. But with something as multi-faceted as political ideology, there is always some risk of residual confounding. My proposed method of building a joint measurement and theory-testing model fixes this issue by obviating the need to deal directly with $U$ at all. This is because the measurement error, $e^\theta$ in @fig-ideal-dag-theory-c is part of the backdoor path from $\bar{\theta}$ to $Y$. Therefore when we explicitly incorporate $e^\theta$ into a model estimating $\bar{\theta} \longrightarrow Y$ we can obtain an unbiased (or at least less-biased, given unobserved measurement error) estimate of the causal effect of ideology on the theoretical outcome of interest.

@@

After the data are generated, I fit two models: the ordinary linear regression using only $\bar{\theta}$ values as was used previously in the attenuation bias example (@eq-cont-no-me-model), and a modified version of @eq-cont-me-model which substitutes the Normal distribution for the Skew Normal distribution (@eq-cont-skew-model). As before, the key parameter of theoretical interest is $\beta_1$---which, according to the simulated data, should equal zero if unconfounded.

@@

I also split the data by party to make the main effects easier to interpret. Lower values from the IRT ideology measurement model correspond to more left-wing candidates, whereas higher values correspond to more right-wing candidates. So, for Democrats we interpret a *negative* relationship between ideology and vote share as district electorates favoring extremist candidates, whereas a *positive* relationship for Republican candidates would mean district electorates favor more extremists.

@@

@@

@fig-ree-betas shows the posterior distributions for $\beta_1$ in the above models (which represents $\bar{\theta} \longrightarrow Y$: measured candidate ideology's effect on vote share in the general election). All the results point towards both Democrats and Republicans favoring more ideologically extreme candidates. The negative coefficient values in the top panel for Democrats suggest that more left-wing candidates perform better, and the positive coefficients in the bottom panel suggest that more right-wing Republican candidates perform better. In this example, the different models only disagree about the magnitudes and uncertainty of these effects. The mean-only models have coefficient values closer to zero compared to the measurement error models. This points to some amount of attenuation bias taking place. Furthermore, the skew-measurement error models show larger electoral effects for extremist candidates, which could mean that there is some minor confounding in @fig-ree that the other models are not taking into account.

@@

In the same vein as using the best computational sampling methods, this project highlights why latent variables whose measurement models produce rich uncertainty estimates should be preferred over those that do not. The methodological competitor to IRT models for measuring political ideology is DW-NOMINATE, which uses multidimensional scaling rather than Bayesian estimation. This optimization method does not provide explicit estimates of uncertainty, much less a full posterior distribution of plausible values the latent variable could take. All of DW-NOMINATE's uncertainty estimates must come from bootstrap procedures which produce, at best, only standard error estimates. This means that more of this model's measurement error is in the unobserved category, which in turn raises concerns about attenuation bias and/or confounding bias when using DW-NOMINATE values in a theory-testing model.
